---
title: "Diversity"
author: "Tianpei Qian"
date: "2023-01-18"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

### TL;DR

- Sign on bonus is given out equally across gender and degree level but very differently across departments. HR department has the lowest bonus rate while the engineering department has the highest.

- The HR department hire much more females than males while we observe the opposite in all the other departments.

- A female is less likely to be on the managing level compared with a male in the sample department, with the same degree level and same years of experience! The problem seems to even get worse at higher levels.

- Salary-wise, the engineering department has the highest salary while the HR department has the lowest. We don't see gender inequality in salary.

### Step 1: read data 

```{r}
company_hierarchy_file <- 'company_hierarchy.csv'
employee_file <- 'employee.csv'

company_hierarchy <- 
  read_csv(
    company_hierarchy_file,
    col_types = list(col_integer(), col_integer(), col_character()) 
  )
```

```{r}
employee <- 
  read_csv(
    employee_file,
    col_types = 
      list(
        col_integer(), 
        col_integer(), 
        col_double(), 
        col_character(), 
        col_character(),
        col_double()
      ) 
  )
```


### Step 2: data cleaning

Join tables, process factors, etc.
```{r}
employee_combined <- 
  employee %>% 
  inner_join(company_hierarchy, by = c("employee_id" = "employee_id")) %>% 
  mutate(
    sex = factor(sex, levels = c("M", "F")),
    degree_level = factor(degree_level, levels = c("High_School","Bachelor","Master","PhD")),
    dept = factor(dept, levels = c("engineering","HR","marketing","sales", "CEO"))
  )
```

### Step 3: quick check

Mainly check on 1-dimensional data. Look out for outliers/missing values.

```{r}
employee_combined %>% 
  summary()
```

First, CEO is a special department with only 1 data point. It probably also contributes the only missing data for boss_id.

```{r}
employee_combined %>% 
  filter(dept == 'CEO')
```

We may or may not remove this CEO data point. It won't cause a big problem if we keep it and handle it with care.

It looks like years of experience may have outliers.

```{r}
employee_combined %>% 
  ggplot(aes(yrs_experience)) +
  geom_boxplot()
```

Most employees have fewer than 10 years of expereince in the company.

### Step 4: data transformation

Now that we're familiar with the data, let's derive some additional variables. Specifically, let's figure out the level of each employee and how many people they manage.

```{r}
rank_levels <- c("IC", "MM", "D", "VP", "E", "CEO")
```

The following is in fact a wrong approach. It assumes that the tree is complete while there is a director that doesn't manage anyone!
```{r eval = FALSE}
employee_ranks_processed <- 
  employee_combined %>% 
  left_join(employee_combined, by = c("employee_id" = "boss_id"), suffix = c("", "_sub")) %>% 
  filter(is.na(employee_id_sub)) %>% 

  
  transmute(
    employee_id,
    boss_id,
    num_reports = 0,
    rank = factor("IC", levels = rank_levels)
  )

employee_ranks <- employee_ranks_processed

for (rank_level in rank_levels[2:length(rank_levels)]) {
  employee_ranks_processed <- 
    employee_combined %>% 
    inner_join(
      employee_ranks_processed, by = c("employee_id" = "boss_id"), suffix = c("", "_sub")
    ) %>% 
    group_by(employee_id) %>% 
    summarize(
      boss_id = max(boss_id),
      num_reports = sum(num_reports) + n()
    ) %>% 
    mutate(
      rank = factor(rank_level, levels = rank_levels)
    )
  employee_ranks <- employee_ranks %>% rbind(employee_ranks_processed)
}
```


To get the rank, we have to start from the root of the tree, i.e. CEO.

```{r}
employee_ranks <- 
  employee_combined %>% 
  mutate(
    rank = if_else(is.na(boss_id), "CEO", "to be updated")
  )

for (i in 1:(length(rank_levels) - 1)) {
  employee_ranks <- 
    employee_ranks %>% 
    left_join(
      employee_ranks, 
      by = c("boss_id" = "employee_id"), 
      suffix = c("", "_boss")
    ) %>% 
    mutate(
      rank = 
        if_else(
          rank_boss == rank_levels[length(rank_levels) - i + 1], 
          rank_levels[length(rank_levels) - i], 
          rank,
          rank # be careful! rank_boss can be null
        )
    ) %>% 
    select_at(vars(-contains("_boss")))
}
```

Next we need to start from the leaf of the tree to get the number of reports for each employee.

```{r}
employee_reports <- 
  employee_ranks %>% 
  filter(rank == rank_levels[1]) %>% 
  transmute(
    employee_id,
    boss_id,
    num_reports = 0
  )

for (rank_level in rank_levels[-1]) {
  employee_reports <- 
    employee_ranks %>% 
    filter(rank == rank_level) %>% 
    left_join(employee_reports, by = c("employee_id" = "boss_id"), suffix = c("", "_sub")) %>% 
    group_by(employee_id, boss_id) %>% 
    summarize(
      num_reports = sum(num_reports) + n(),
      # suffix only comes into play when there's duplicate!!
      .groups = "drop"
    ) %>% 
    mutate(num_reports = if_else(is.na(num_reports), 0, num_reports)) %>% 
    rbind(employee_reports)
}
```

Combine everything and take a final look
```{r}
employee_final <- 
  employee_ranks %>% 
  inner_join(employee_reports, by = c("employee_id" = "employee_id", "boss_id" = "boss_id")) %>% 
  filter(rank != "CEO") %>% 
  mutate(rank = factor(rank, levels = rank_levels))

employee_final %>% summary()
```

Another approach is to use recursion, which is capcable of traversing the tree top-down and bottom-up at one go.
```{r eval = FALSE}
employee_combined_re <- 
  employee_combined %>% 
  mutate(
    boss_id = if_else(is.na(boss_id), -1L, boss_id),
    rank = as.character(NA),
    num_reports = as.integer(NA)
  )

find_rank <- function(rank_idx, id) {
  employee_combined_re$rank[employee_combined_re$employee_id == id] <<- rank_levels[rank_idx]
    
  if (rank_idx == 1) {
    employee_combined_re$num_reports[employee_combined_re$employee_id == id] <<- 0
    return(0)
  }
  
  ct <- 0
  for (id_sub in employee_combined_re$employee_id[employee_combined_re$boss_id == id] ) {
    ct <- ct + find_rank(rank_idx - 1, id_sub) + 1
  }
  employee_combined_re$num_reports[employee_combined_re$employee_id == id] <<- ct
  return(ct)
}

find_rank(
  length(rank_levels), 
  employee_combined_re %>% filter(boss_id == -1) %>% pull(employee_id)
)
```



### Step 4: exploratory analysis

Now we're finally ready to answer the million dollar question: do you think the company has been treating all its employees fairly?


#### Sign on bonus
Is sign on bonus given out fairly?

```{r}
employee_final %>% 
  group_by(dept, degree_level, sex) %>% 
  summarize(
    bonus_rate = sum(signing_bonus) / n()
  ) %>% 
  ggplot(aes(dept, bonus_rate)) +
  geom_line(aes(color = degree_level, group = degree_level)) +
  facet_grid(vars(sex))
```

A quick plot shows that sex and degree level seems irrelevant to bonus rate but it seems to be lower in the HR department. There is no obvious interaction pattern between these variables. 

```{r}
bonus_lm <- 
  glm(
    signing_bonus ~ degree_level + sex + dept, 
    data = employee_final, 
    family = 'binomial'
  )
```

```{r}
bonus_lm %>% summary()
```

Indeed, department has a significant impact on bonus rates, with the HR department being the lowest and the engineering department being the highest.


Note that in answering this question we shouldn't consider more variables than needed!! This is a common pitfall that people new to modelling can fall into. For example, gender may affect the level the person gets and the level can subsequently decide whether the person gets sign on bonus. If you include "level" in your model when studying the impact of gender, then you will get an inaccurate estimate because you limit the impact of gender on level.


#### Department

Does each department have the same criteria for hiring?

```{r}
employee_final %>% 
  group_by(dept, degree_level) %>% 
  summarize(
    female_rate = sum(sex == 'F') / n()
  ) %>% 
  ggplot(aes(dept, female_rate)) +
  geom_line(aes(color = degree_level, group = degree_level))
```

The HR department hires more females than males while the other 3 departments hire more males.

A chi-square test confirms that each department has significantly different gender distributions.

```{r}
employee_final %>% 
  count(sex, dept) %>% 
  pivot_wider(names_from = sex, values_from = n) %>% 
  select(-dept) %>% 
  chisq.test()
```

```{r}
employee_final %>% 
  group_by(dept, sex) %>% 
  summarize(
    high_school = sum(degree_level == 'High_School') / n(),
    bachelor = sum(degree_level == 'Bachelor') / n(),
    master = sum(degree_level == 'Master') / n(),
    phd = sum(degree_level == 'PhD') / n()
  ) %>% 
  pivot_longer(cols = high_school:phd, names_to = "degree", values_to = "prop") %>% 
  ggplot() +
  geom_line(aes(dept, prop, color = degree, group = degree)) +
  facet_grid(cols = vars(sex))
```
On the other hand, all departments seem to have very similar hiring standards on degree levels.


#### Company level

Does female have a fair representation on the managing levels?

```{r}
employee_final %>% 
  group_by(rank) %>% 
  summarize(
    female_rate = sum(sex == "F") / n()
  ) %>% 
  ggplot(aes(rank, female_rate)) +
  geom_col()
```

Firstly, the company has fewer females across all levels and female representation seems to get worse as the level increases.

To get a more rigorous analysis, let's build a model that also controls department, degree level and years of experience.

```{r}
manager_lm <- 
  glm(
    is_manager ~ degree_level + sex + dept + yrs_experience, 
    data = employee_final %>% mutate(is_manager = (rank != "IC")), 
    family = 'binomial'
  )
```

```{r}
manager_lm %>% summary()
```

Not surprisingly, you have roughly equal opportunities to become managers in all departments. Also, higher degrees and more years of experience increase the odds. Lastly and notably, females are less likely to become managers in this company even when they're in the same department, have the same degree level and have the same years of experience!

#### Salary

Finally, do we see inequality in salary?

```{r}
employee_final %>% 
  group_by(dept, rank, sex) %>% 
  summarize(salary = mean(salary)) %>% 
  ggplot(aes(dept, salary)) +
  geom_point(aes(color = rank, group = rank)) +
  geom_line(aes(color = rank, group = rank)) +
  facet_grid(cols = vars(sex))
```


```{r}
salary_lm <- 
  lm(
    salary ~ degree_level + sex + rank + dept  + yrs_experience + num_reports, 
    data = employee_final
  )
```

```{r}
salary_lm %>% summary()
```

The department and company level matters for salary, while degree level, sex, years of experience and number of reports do not have a significant impact on salary for a given department and company level.

